---
title: "Project 1"
format: pdf
editor: visual
---

# STAT40850 Bayesian Analysis

Packages

```{r}
suppressMessages(library(rstan))
suppressMessages(library(bayesrules))
suppressMessages(library(bayesplot))
```

## Questions :

### 1. Specify and plot a Beta prior distribution P($\theta_A$) with mean E($\theta_A$)=0.65. Calculate the prior probability P($\theta_A$ \< 5).

The **Beta distribution** is parameterised by alpha and beta, where:

E($\theta_A$) = 0.65 = alpha / alpha + beta

```{r}
#since E(theta_A) = 0.65, we take alpha = 6.5 and beta = 3.5 
alpha = 6.5 
beta = 3.5  
#ploting the graph 
curve(dbeta(x, alpha, beta), col = 'blue', yaxt = 'n',        
      lwd = 3, main = 'Beta prior distribution',       
      xlab = expression(theta[A]), ylab = 'Density',        
      from = 0, to = 1)
```

```{r}
#Computing p(theta_A < 5) 
prior_prob = pbeta(0.5, alpha, beta) 
cat("Prior probability less than 0.5 is", prior_prob)
```

### 2. Estimate and plot the posterior distribution p($\theta_A$ ∣ xA). Calculate the posterior probability Pr($\theta_A$ \> 0.7).

-   **For posterior distribution, theta_A ​∣ xA​ ∼ Beta( alpha + xA ​, beta + nA ​− xA ​)**

```{r}
# Updating posterior parameters 
alpha_post = alpha + 169 
beta_post = beta + (296 - 169)  
#plotting the graph for the updated posterior parameters 
curve(dbeta(x, alpha_post, beta_post), col = 'blue', yaxt = 'n',        
      lwd = 3, main = 'Beta posterior distribution',       
      xlab = expression(theta[A]), ylab = 'Density',        
      from = 0, to = 1)
```

```{r}
#Computing p(theta_A > 7) 
post_prob = 1 - pbeta(0.7, alpha_post, beta_post) 
cat("Posterior probability of theta_a greater than 0.7 is", post_prob)
```

### 3. Present a brief commentary on the results obtained focusing on whether there is any evidence against the hypothesis that $\theta_A$ = 0.60 a posteriorir.

Let

Null Hypothesis: H0: $\theta_A$ = 0.60

Alternative Hypothesis: Ha: $\theta_A$ != 0.60

```{r}
low_q <- qbeta(0.025, alpha_post, beta_post)
upper_q <- qbeta(0.975, alpha_post, beta_post)
c(low_q, upper_q)
```

-   The **null hypothesis** (H0​) states that $\theta_A$ = 0.60, while the **alternative hypothesis** (Ha​) states that $\theta_A$ != 0.60.

-   The **95% credible interval** for $\theta_A$​, calculated using the quantile method, is **\[0.5178, 0.6284\]**.

-   Since **0.60 lies within this interval**, we **fail to reject the null hypothesis**.

-   There is **insufficient evidence** to conclude that $\theta_A$ !=0.60.

### 4. Estimate (via Monte Carlo sampling) and plot the posterior predictive distribution Pr(xA∣xA). Comment on the fit of the model to the observed data. Use the sample simulated from the posterior predictive distribution to estimate the probability Pr(xA \>=180 ∣ xA).

```{r}
#Number of simulations 
n_sim = 300  
set.seed(1)  
#Estimating theta value 
theta_s = rbeta(n_sim, alpha_post, beta_post)  
#Simulating future student pass counts using a Binomial model 
x_a = rbinom(n_sim, 296, theta_s)  
#Plotting 
hist(x_a, breaks = 30, col = "red", 
     main = 'Posterior Predictive Distribution', 
     xlab = 'Number of Students Passing', 
     freq = FALSE, border="black")
```

```{r}
cat("Median Xa: ", median(x_a))
```

**Comment on the Fit of the Model:**

-   The distribution appears **bell-shaped**, centered around **160–170 passes**.

-   The actual observed value (169) is **consistent** with the predicted range.

-   Median 169 suggests that **our model fits the observed data well**.

```{r}
prob_xA_180 = mean(x_a >= 180) 
cat("Probability of x_A >= 180 given x_A is", prob_xA_180)
```

**Interpretation:**

The probability of 180+ student passing in the exam is **19.7%.** This align well with the expectation.

### 5.Use **Stan** to estimate the posterior distribution for the probability of passing the exam for each group ($\theta_A$ and $\theta_B$) using the same prior adopted above for $\theta_A$ and a symmetric prior for $\theta_B$. Estimate the posterior distribution of the difference $\theta_diff$= $\theta_B$ − $\theta_A$.

```{r}
stan_code = ' data {   
int<lower=0> n_A;  // Total students in group A   
int<lower=0> y_A;  // Students passing in group A   
int<lower=0> n_B;  // Total students in group B   
int<lower=0> y_B;  // Students passing in group B   
real<lower=0> alpha_A; // Prior alpha for θA   
real<lower=0> beta_A;  // Prior beta for θA   
real<lower=0> alpha_B; // Prior alpha for θB   
real<lower=0> beta_B;  // Prior beta for θB 
}  
parameters {   
// Probability of passing for Group A   
real<lower=0, upper=1> theta_A;     
// Probability of passing for Group B   
real<lower=0, upper=1> theta_B;   
}  
transformed parameters {   
// Difference in pass rates   
real theta_diff = theta_B - theta_A;   
}  
model {   
// Priors   
theta_A ~ beta(alpha_A, beta_A);   
theta_B ~ beta(alpha_B, beta_B);    
// Likelihood   
y_A ~ binomial(n_A, theta_A);   
y_B ~ binomial(n_B, theta_B); 
}  
generated quantities {   
real theta_diff_post = theta_B - theta_A; 
} '
```

```{r}
#Replacing data  
data_list <- list(   
  n_A = 296,   
  y_A = 169,   
  n_B = 380,   
  y_B = 247,   
  alpha_A = 6.5, 
  beta_A = 3.5,   
  alpha_B = 5, 
  beta_B = 5 
  )  
#Fitting Stan model 
st_model <- stan_model(model_code = stan_code)
fit = sampling(st_model,
       data = data_list,              
       iter = 4000, 
       seed = 1,
       refresh = 0
       )  
#summary statistics 
print(fit, pars = c("theta_A", "theta_B", "theta_diff_post")) 
```

### 6. Plot and summarize the posterior distributions estimated in the previous question; estimate of the posterior probability Pr($\theta_diff$ \< 0∣xA,xB) by using the Stan MCMC output; present a brief commentary on the results obtained.

```{r}
posterior_samples = as.data.frame(fit)
mcmc_hist(posterior_samples, pars = c('theta_A','theta_B','theta_diff'))
```

**Estimating Pr(**$\theta_diff$**\< 0 \| xA, xB):**

To estimate this probability, we need to count how many posterior samples of $\theta_diff$are negative and divide it by the total number of samples.

```{r}
posterior_sample_extract = extract(fit)
prob_theta_diff_neg <- mean(posterior_sample_extract$theta_diff < 0)
cat("Posterior probability Pr(theta_diff < 0 | xA, xB):", prob_theta_diff_neg)
```

**Interpretation:**

The **posterior probability** P(theta_diff \< 0) is **2.34%**, indicating a very low chance that **Method B performs worse than Method A**.

-   Since this probability is **below 5%**, there is **strong evidence** that **Method B has a higher pass rate** than Method A.

-   If P(theta_diff \< 0) were **around 50%**, both methods would be similar, but the low value strongly **favors Method B**.

-   This suggests that **Method B is the better choice** for improving student performance.
